Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-04-05 analysis/paper_draft2.Rmd
Update paper_draft2.Rmd
Tim and Eric edits to intro and methods

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018a; @Kadane2018b] to incorporate historical strike data. In this approach, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. Then, we use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. In so doing, we use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter.  Finally, we use that posterior distribution as the prior for the bias parameter in the present case and update the posterior based on data on the target attorney's strikes in that present case.

@@

We conducted a simulation study to measure model performance under different values of the bias parameter $b$ and different amounts and kinds of historical information.

To do this, we generated strike data for a current trial using seven different values of the bias parameter ($b_{curr} = \{-3,-2,-1,0,1,2,3\}$. As introduced above, $b>0$ ($b<0$) represents bias against (for) a prospective juror within (outside) the cognizable class, while $b=0$ denotes no bias.

We also generated historical strike data of three sizes (same, double, and triple the size of current data, i.e., data on one previous trial, two previous trials) using seven different values of the bias parameter for generating that historical data ($b_{hist} = \{-3,-2,-1,0,1,2,3\}$).

We let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) to simulate noise associated with historical strike data that is incomplete not at random. Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature of the current trial but not of all the past trials in the historical data. To illustrate, suppose prosecutor with a bias value conditional on defendant race: $b > 0$ if the criminal defendant is Black, $b = 0$ if not. If our historical strike data contains only in cases with non-Black defendants, accounting for that historical data may lead us to underestimate prosecutor's bias in the current case ($b_{curr}$) with a Black defendant.

Because we use the power prior, we can control how much we account for the historical strike data by modifying the weight parameter $a$. If $a = 1$, the historical strike data is equally weighted with the data on strikes in the current trial. If $a < 1$, the historical data matters proportionally less. We can assign $a < 1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we tried different values of the weight parameter ($a = \{0.1,0.3,0.5,0.7,1\}$) to show how that parameter affected modeling accuracy.

We considered 735 scenarios in total, i.e., the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data and 5 different values of the power prior weight parameter $a$. For each scenario, we generated 1000 datasets and fit the model on those datasets. We recorded the posterior mean and credible intervals of one thousand model fits for each scenario.

In turn, for three credible interval levels (80\%, 90\% and 95\%), we recorded the bias detection rates, i.e., the proportion of times we can reach the correct conclusion about the bias parameter based on those credible intervals. That is, when the true value of $b_{curr}$ is greater than zero, the credible interval lies to the right of zero; when $b_{curr}$ equals to zero, the credible interval includes zero; and when $b_{curr}$ is less than zero, the credible interval does not include zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus _not_ on how accurately the model can recover the true value of $b_{curr}$ (i.e., a traditional coverage rate), but on how accurately our model can detect that bias. For instance, a 90\% bias detection rate means that in 900 of 1000 model fits, a given credible interval does not contain zero if the $b_{curr}$ is not zero. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero (lower bound is positive) among the 1000 model fits.

@@

Figures 1-3 depicts the coverage rates of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively, for the combinations of $b_{hist}$, $b_{curr}$, the power prior weight parameter $a$, and different sizes of the historical data ($h = {1,2,3}$). For any given combination, a darker color shade denotes a higher coverage rate, and vice versa.

>>>>

In this paper, we extend a Bayesian approach to _Batson_ and similar challenges [@Kadane2018a; @Kadane2018b] to incorporate historical strike data. In this approach, we use two steps. In step one, we specify a model of the peremptory-strike process in the court of interest that includes a bias parameter to which we assign an initial prior distribution. We use an attorney's strike data from past trials, as generated by that same peremptory-strike process, to estimate a posterior distribution for that bias parameter. We use the power prior [@Chen2000; @Ibrahim2015] to control how much the strike history from past cases affects the posterior distribution of the bias parameter. In step two, we estimate the posterior distribution of the bias parameter from the current case, using the posterior distribution developed with the historical information and the power prior.

@@

To evaluate the performance of the proposed method, we conducted a simulation study. The primary aim of these simulations is to examine the effect of the strength of discounting of the historical data by the power prior ($\alpha$), and the size of the historical data set, on our ability to detect bias.

Because we use the power prior, we can control how much we account for the historical strike
data by modifying the discounting parameter $\alpha$. If $\alpha=1$, the historical strike data is equally
weighted with the data on strikes in the current trial. If $\alpha<1$, the historical data are discounted and weighted proportionally less than the current trial data. We assign $\alpha<1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we evaluated different values of $\alpha$: $\alpha=\{0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1\}$ to show how that parameter affected bias detection .

We generated historical strike data of three sizes (same, double, and triple the size of
current data, i.e., data on one previous trial, two previous trials) using seven different values
of the bias parameter for generating that historical data $b_{hist}=\{-3,-2,-1,0,1,2,3\}$.

To generate data, we generated strike data for a current trial using seven different values of the bias parameter $b_{curr}=\{0,0.5,1,1.5,2, 2.5,3\}$. These values were selected to evaluate the strength of the power prior as a function of the compatibility between current and historical trial data. As introduced above, $b > 0$ ($b < 0$) represents bias against (for) a prospective juror within (outside) the cognizable class, while $b = 0$ denotes no bias.

We let $b_{hist}$ differ from $b_{curr}$ (our parameter of interest) to simulate noise associated with
historical strike data that is incomplete not at random, by allowing $b_{hist}$ to have the same selected bias values, i.e., $b_{hist}=\{0,0.5,1,1.5,2, 2.5,3\}$ . Alternatively, this difference simulates a real difference in the bias parameter that depends on a feature of the current trial but
not of all the past trials in the historical data. To illustrate, suppose prosecutor with a bias
value conditional on defendant race: $b > 0$ if the criminal defendant is Black, $b = 0$ if not.
If our historical strike data contains only cases with non-Black defendants, accounting
for that historical data may lead us to underestimate prosecutor’s bias in the current case
$b_{curr}$ with a Black defendant. We note that we assume the defense has no bias.

For our simulation we define bias detection as the proportion of times we identify bias when bias is present in the current trial, i.e., $\b_{curr} > 0$ , based on whether credible intervals of the bias parameter exclude zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus _not_ on how accurately the model can recover the true value of $b_{curr}$ (i.e., a traditional coverage rate), but on how accurately our model can detect that bias. For instance, a 90\% bias detection rate means that in 900 of 1000 model fits, a given credible interval does not contain zero if the $b_{curr}$ is not zero. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero (lower bound is positive) among the 1000 model fits. For each scenario, we generated 1000 data sets and fit the model on those data sets. For each replicate, we generate a MCMC sample of 10000 iterations with a burn-in period of 1000 iterations. For each replicate, we calculate the posterior mean and $80\%$, $90\%$, and $95\%$ highest posterior density (HPD) intervals for $\b_{curr}$.

Additionally, we present sensitivity analysis for the number of available strikes with 6 strikes, 10 strikes, and 15 strikes. Thus, in total, we considered 4851 different scenarios, i.e. the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data, 11 different values for the power prior weight parameter, and for 3 combinations of the total number of strikes.

@@

Figures 1-3 depicts the bias detection of $b_{curr}$ with an 80\%, 90\%, and 95\% credible interval, respectively, for the 7 combinations of $b_{hist}$, and 7 combinations of $b_{curr}$, three different levels the power prior discounting parameter $a$, and different sizes of the historical data ($h = {1,2,3}$).
